Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-01 paper/main.tex
final edits to paper

Over the past several years, the global pandemic has forced numerous educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
shared and viewed by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized and/or augmented by automated tools? How and why do
learning needs and goals vary across people? How might we lower barriers to
achieving a high-quality education?

@@

We enrolled a total of 50 Dartmouth undergraduate students in our study.
Participants received course credit for enrolling. We asked each participant to
fill out a demographic survey that included questions about their age, gender,
native spoken language, ethnicity, race, hearing, color vision, sleep, coffee
consumption, level of alertness, and several aspects of their educational
background and prior coursework.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
semantically related or that tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic. (Unless, otherwise noted, we used $k = 15$
topics.)

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual transcriptions
of all videos for closed captioning. However, such transcripts would not be
readily available in all contexts to which our framework could potentially be
applied. Khan Academy videos are hosted on the YouTube platform, which
additionally provides automated captions. We opted to use these automated
transcripts (which, in prior work, we have found are sufficiently near-human
quality yield reliable data in behavioral studies; \citealp{ZimaEtal18}) when
developing our framework in order to make it more directly extensible and
adaptable by others in the future.

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo19}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). We defined a sliding window
length of (up to) $w = 30$ transcript lines, and assigned each window a
timestamp corresponding to the midpoint between its first and last lines'
timestamps. These sliding windows ramped up and down in length at the very
beginning and end of the transcript, respectively. In other words, the first
sliding window covered only the first line from the transcript; the second
sliding window covered the first two lines; and so on. This insured that each
line of the transcript appeared in the same number ($w$) of sliding windows.
After performing various standard text preprocessing (e.g., normalizing case,
lemmatizing, removing punctuation and stop-words), we treated the text from
each sliding window as a single ``document,'' and we combined these documents
across the two videos' windows to create a single training corpus for the topic
model. The top words from each of the 15 discovered topics may be found in
Table~\topics.

After fitting a topic model to each videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of Euclidean distance, correlation, or other geometric measures). In
general, the similarity between different documents' topic vectors may be used
to characterize the similarity in conceptual content between the documents.

@@

\begin{equation}
    \mathrm{ncorr}(x, y) = \frac{\mathrm{corr}(x, y) - \mathrm{mincorr}}{\mathrm{maxcorr} - \mathrm{mincorr}},
\end{equation}
and where $\mathrm{mincorr}$ and $\mathrm{maxcorr}$ are the minimum and maximum
correlations between any lecture timepoint and question, taken over all
timepoints and questions across both lectures and all five question used to
estimate the knowledge trace. We also define $f(s, \Omega)$ as the
$s$\textsuperscript{th} topic vector from the set of topic vectors $\Omega$.
Here $t$ indexes the set of lecture topic vectors, $L$, and $i$ and $j$ index
the topic vectors of questions used to estimate the knowledge trace, $Q$. Note
that ``correct'' denotes the set of indices of the questions the participant
answered correctly on the given quiz.

@@

At a high level, the UMAP algorithm obtains low-dimensional embeddings by
minimizing the cross entropy between the pairwise (clustered) distances between
the observations in their original (e.g., 100-dimensional) space and the
pairwise (clustered) distances in the low-dimensional embedding space (in our
approach, the embedding space is 2D). In our implementation, pairwise distances
in the original high-dimensional space were defined as 1 minus the correlation
between the pair of coordinates, and pairwise distances in the low-dimensional
embedding space were defiend as the Euclidean distance between the pair of
coordinates.

@@

We acknowledge useful discussions, assistance in setting up an earlier
(unpublished) version of this study, and assistance with some of the data
collection efforts from Will Baxley, Max Bluestone, Daniel Carstensen, Kunal
Jha, Caroline Lee, Lucy Owen, Xinming Xu, and Kirsten Ziman. Our work was
supported in part by NSF CAREER Award Number 2145172 to JRM. The content is
solely the responsibility of the authors and does not necessarily represent the
official views of our supporting organizations. The funders had no role in
study design, data collection and analysis, decision to publish, or preparation
of the manuscript.

>>>>

Over the past several years, the global pandemic has forced many educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
shared and viewed by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized and/or augmented by automated tools? How and why do
learning needs and goals vary across people? How might we lower barriers to
achieving a high-quality education?

@@

We enrolled a total of 50 Dartmouth undergraduate students in our study.
Participants received course credit for enrolling. We asked each participant to
complete a demographic survey that included questions about their age, gender,
native spoken language, ethnicity, race, hearing, color vision, sleep, coffee
consumption, level of alertness, and several aspects of their educational
background and prior coursework.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
semantically related or tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic. (Unless, otherwise noted, we used $k = 15$
topics.)

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual transcriptions
of all videos for closed captioning. However, such transcripts would not be
readily available in all contexts to which our framework could potentially be
applied. Khan Academy videos are hosted on the YouTube platform, which
additionally provides automated captions. We opted to use these automated
transcripts (which, in prior work, we have found are of sufficiently near-human
quality yield reliable data in behavioral studies; \citealp{ZimaEtal18}) when
developing our framework in order to make it more directly extensible and
adaptable by others in the future.

We fetched these automated transcripts using the
\texttt{youtube-transcript-api} Python package~\citep{Depo19}. The transcripts
consisted of one timestamped line of text for every few seconds (mean: 2.34~s;
standard deviation: 0.83~s) of spoken content in the video (i.e., corresponding
to each individual caption that would appear on-screen if viewing the lecture
via YouTube, and when those lines would appear). We defined a sliding window
length of (up to) $w = 30$ transcript lines, and assigned each window a
timestamp corresponding to the midpoint between its first and last lines'
timestamps. These sliding windows ramped up and down in length at the very
beginning and end of the transcript, respectively. In other words, the first
sliding window covered only the first line from the transcript; the second
sliding window covered the first two lines; and so on. This insured that each
line of the transcript appeared in the same number ($w$) of sliding windows.
After performing various standard text preprocessing (e.g., normalizing case,
lemmatizing, removing punctuation and stop-words), we treated the text from
each sliding window as a single ``document,'' and combined these documents
across the two videos' windows to create a single training corpus for the topic
model. The top words from each of the 15 discovered topics may be found in
Table~\topics.

After fitting a topic model to the two videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of Euclidean distance, correlation, or other geometric measures). In
general, the similarity between different documents' topic vectors may be used
to characterize the similarity in conceptual content between the documents.

@@

\begin{equation}
    \mathrm{ncorr}(x, y) = \frac{\mathrm{corr}(x, y) - \mathrm{mincorr}}{\mathrm{maxcorr} - \mathrm{mincorr}},
\end{equation}
and where $\mathrm{mincorr}$ and $\mathrm{maxcorr}$ are the minimum and maximum
correlations between any lecture timepoint and question, taken over all
timepoints in the given lecture, and all five questions \textbf{about} that lecture appearing on the given quiz. 
We also define $f(s, \Omega)$ as the
$s$\textsuperscript{th} topic vector from the set of topic vectors $\Omega$.
Here $t$ indexes the set of lecture topic vectors, $L$, and $i$ and $j$ index
the topic vectors of questions used to estimate the knowledge trace, $Q$. Note
that ``correct'' denotes the set of indices of the questions the participant
answered correctly on the given quiz.

@@

At a high level, the UMAP algorithm obtains low-dimensional embeddings by
minimizing the cross-entropy between the pairwise (clustered) distances between
the observations in their original (e.g., 100-dimensional) space and the
pairwise (clustered) distances in the low-dimensional embedding space (in our
approach, the embedding space is 2D). In our implementation, pairwise distances
in the original high-dimensional space were defined as 1 minus the correlation
between the pair of coordinates, and pairwise distances in the low-dimensional
embedding space were defiend as the Euclidean distance between the pair of
coordinates.

@@

We acknowledge useful discussions, assistance in setting up an earlier
(unpublished) version of this study, and assistance with data
collection efforts from Will Baxley, Max Bluestone, Daniel Carstensen, Kunal
Jha, Caroline Lee, Lucy Owen, Xinming Xu, and Kirsten Ziman. Our work was
supported in part by NSF CAREER Award Number 2145172 to JRM. The content is
solely the responsibility of the authors and does not necessarily represent the
official views of our supporting organizations. The funders had no role in
study design, data collection and analysis, decision to publish, or preparation
of the manuscript.
